Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/study.tex
various edits, shrinking, and conclusion
section 1 and 2 edits/nits

\item Navier, a small-size Navier-Stokes fluid model (based
  on \cite{griebel1997numerical};

@@

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we classify the relationship between the LHS
and RHS in terms of the use of induction variables.
The two sides are \emph{directly consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the LHS
induction variables are either
a subset or superset of the induction variables on the RHS.
Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l|l}
  %Shorthand &
  Classification & Sub Classification & Example \\ \hline
  %(\textsf{const}) &
  Consistent & Direct & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  %(\textsf{perm})
  & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  %(\textsf{LHSsub}) &
  & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  %(\textsf{LHSsup}) &
  & LHS superset & \fortran{a(i, j) = b(i)} \\ \hline
  %(\textsf{inconst}) &
  \multicolumn{2}{l|}{Inconsistent} & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%

@@

\noindent
We revisit each hypothesis and show the related data and conclusions.
%
\begin{enumerate}
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or derived) induction variables;
\end{enumerate}
%
We identified \num{108773} instances of assignments within loops in
which an array subscript flows to the right-hand side. We refer
to each one of these as an \emph{array computation}.
We performed the classification described above
and grouped the data by the right-hand side classification only:
%%
\begin{center}
\begin{tabular}{l|S[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
\textbf{RHS classification} & \textbf{Number} & \% &
\textit{Grouping 1} \\ \hline
Affine                          & 34        & 0.03\%
& \multirow{3}{*}{81.43\%} \\ \cline{1-3}
Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
Neighbourhood + absolute        & 23209     & 21.34\%  \\ \hline
Other                           & 20196     & 18.57 \%  \\ \hline \hline
\textbf{Total}                  & 108773    & 100.00\% \\
\end{tabular}
\end{center}
%
\noindent
Since, affine and neighbour patterns are static, we group them
(\textit{Grouping 1}) to see that $81.43\%$ of the array computations
(88,577) have a static pattern that neighbourhood or affine subscript
expressions (potential with some absolute
indices), confirming \emph{Hypothesis 1}. Note that affine subscripts
are rare, perhaps due to programmers relying on compilers for
optimisation rather than hand optimising (which tends to lead to
affine indices).  Overwhelmingly, the main category is neighbourhood
offsets (possibly mixed with constants in some dimensions). The
``affine + absolute'' class is not represented at all. %\marginpar{\mnote{(2)
%  Mention what sort of things are included in other.}}

\begin{enumerate}[resume]
\item Most loop-array computations of the previous form read
from arrays with a \emph{contiguous} pattern.
%
\item Most loop-array computations of the previous form read
from arrays with a pattern that includes the immediate
neighbours (offsets of 1 from the induction variables);
\end{enumerate}
%
We grouped the data based on subclassifications of \emph{Grouping 1}
in terms of contiguity and positioning.
%Focussing on grouping 1.
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
\textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
&\textbf{Position} & \textbf{Number} & \% {(of whole)} \\ \hline
Affine, contiguous
& 31 & 0.03\% \\ \hline
%\multirow{3}{*}{Affine, contiguous}
% & %\multirow{3}{*}{31}
% & %\multirow{3}{*}{0.03\%}
% & \textsf{away} & 4 & 0.00\% \\
%& 31 & 0.03\% & \textsf{origin} & 15 & 0.01\% \\
%&&& \textsf{straddle} & 12 & 0.01\% \\ \hline
Affine, non-contiguous
 & 3
 & 0.00\% & \\ \hline
\multirow{3}{*}{Neighbour, contiguous}
 & %\multirow{3}{*}{83689}
 & %\multirow{3}{*}{83.08\%}
 & \textsf{away} & 152 & 0.14\% \\
& 83689 & 76.94\% & \textsf{origin} & 81309 & 74.75\% \\
&&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
Neighbour, non-contiguous
 & 4854
 & 4.46\% \\ \hline
Other & 20196 & 18.57\%  \\
 \hline \hline
\textbf{Total}  & 108773    & 100.00\% \\
\end{tabular}
\end{center}
%%
Thus, $76.94\%$ of array computations have only neighbour
indices (which may include some absolute indices) in a contiguous
pattern on the right hand side, with $76.80\%$ either including the origin
($74.75\%$) or offset by a distance of 1 from the origin
($2.05\%$). This confirms both Hypothesis 2 and Hypothesis 3. Only 152 instances
of index patterns not immediately next-to or over the origin were
found.  There is a non trivial amount of non-contiguous arrays access,
but it is a much smaller proportion of the whole.

@@

We subclassified our data by r.h.s
and l.h.s categories

@@

neighbour-indexed or affined-indexed array. We additionally

@@

& \textbf{\# Consistent}

@@

The most common category has an LHS array subscript

@@

Most stencils are contiguous, but there is some
non-contiguity. Most stencils have matching induction
variables on the left and right-hand sides but again there
are non-trivial amounts of the other kinds of relationship.
%%
\begin{enumerate}[resume]
\item Many array computations read from each particular index just
  once.
\end{enumerate}
%
We classified right-hand sides comprising neighbourhood (or
affine) subscripts and subclassified this by whether array subscript terms were
all unique or whether there were some repeats.
%
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  %\\\hline\hline
%Total                             & 108773    &
\end{tabular}
\end{center}
%
Thus most array computations have access patterns with
unique use of array subscripts, confirming Hypothesis 5.

@@

\noindent
For each of the $108,773$ array computations, we recorded four
additional properties: the max dimensionality of array
subscripts involved, %the maximum depth of neighbourhood offsets,
the number of subscript terms involved in the computation, and the
length of the dataflow path (\ie{}, how many intermediate assignments
contribute to the array computation). The following gives the data
as histograms:

@@

 \begin{center}
\begin{tabular}{l|ccccccccccccccccccccc}
  \textbf{Dataflow length}: & 0      & 1     & 2    & 3
  & 4    & 5 & 6    & 7   & 8   & 9   & 10  & 11  & 12  & 13  & 14
  & 15$\leq$ \\\hline
  Count & 106,712 & 13,323 & 7,091 & 3,898 & 3,031 & 1,695 & 1,333 & 965 & 792 & 704 & 482 & 430 & 339 & 310 & 270 & 2544
\end{tabular}
\end{center}
%
The histogram of ``number of subscripts'' shows that most
computations are relatively simple, with only 1-3 array subscripts.
However, some are complex, with 184 with over 20 indexing terms. We had one example with 96
contributing array subscripts. Thus, we see that there are many cases
in practice with complex patterns, that we aim to provide support for
with our specification language.

@@

\item Access patterns are mostly contiguous ($76.94\%$) and cross or
straddle the origin ($75.75\%$), thus we can represent patterns by
\emph{finite intervals} that meet at, or are next to, the origin.
These intervals will be given declaratively and abstractly, avoiding
replicating indexing terms at the specification level.

@@

\item Stencils will be the primary focus $(61.79\%)$ but we will also
allow specifications on assignments where the l.h.s is a variable
(\eg{}, in a reduction) ($4.22\%$).

@@

\item Affine indexing is very rare $<0.03\%$ so we do not consider it.
\end{itemize}
%

>>>>

\item Navier, a small-size Navier-Stokes fluid model, based
  on \cite{griebel1997numerical};

@@

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we classify the relationship between the LHS
and RHS in terms of the use of induction variables.
The two sides are \emph{directly consistent} if the same induction variables
appear in each side, used for the same dimensions. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the LHS
induction variables are either
a subset or superset of the induction variables on the RHS.
Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l|l}
  %Shorthand &
  Classification & Sub Classification & Example \\ \hline
  %(\textsf{const}) &
  Consistent & Direct & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  %(\textsf{perm})
  & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  %(\textsf{LHSsub}) &
  & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  %(\textsf{LHSsup}) &
  & LHS superset & \fortran{a(i, j) = b(i)} \\ \hline
  %(\textsf{inconst}) &
  \multicolumn{2}{l|}{Inconsistent} & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%

@@

\noindent
We revisit each hypothesis and show the related data and conclusions.
%
\begin{enumerate}
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or derived) induction variables;
\end{enumerate}
%
We identified \num{108773} instances of assignments within loops in
which an array subscript flows to the right-hand side. We refer
to each one of these as an \emph{array computation}.
We performed the classification described above
and grouped the data by the right-hand side classification only:
%%
\begin{center}
\begin{tabular}{l|S[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
\textbf{RHS classification} & \textbf{Number} & \% &
\textit{Grouping 1} \\ \hline
Affine                          & 34        & 0.03\%
& \multirow{3}{*}{81.43\%} \\ \cline{1-3}
Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
Neighbourhood + absolute        & 23209     & 21.34\%  \\ \hline
Other                           & 20196     & 18.57 \%  \\ \hline \hline
\textbf{Total}                  & 108773    & 100.00\% \\
\end{tabular}
\end{center}
%
\noindent
Since, affine and neighbour patterns are static, we group them
(\textit{Grouping 1}) to see that $81.43\%$ of the array computations
(88,577) have a static pattern comprising neighbourhood or affine subscript
expressions (potentially with some absolute
indices), confirming \emph{Hypothesis 1}. Note that affine subscripts
are rare, perhaps due to programmers relying on compilers for
optimisation rather than hand optimising (which tends to lead to
affine indices).  Overwhelmingly, the main category is neighbourhood
offsets (possibly mixed with absolute indices in some dimensions). The
``affine + absolute'' class is not represented at all. %\marginpar{\mnote{(2)
%  Mention what sort of things are included in other.}}

\begin{enumerate}[resume]
\item Most loop-array computations of the previous form read
from arrays with a \emph{contiguous} pattern.
%
\item Most loop-array computations of the previous form read
from arrays with a pattern that includes the immediate
neighbours (offsets of 1 from the induction variables);
\end{enumerate}
%
We grouped the data based on subclassifications of \emph{Grouping 1}
in terms of contiguity and positioning.
%Focussing on grouping 1.
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
\textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
&\textbf{Position} & \textbf{Number} & \% {(of whole)} \\ \hline
Affine, contiguous
& 31 & 0.03\% \\ \hline
%\multirow{3}{*}{Affine, contiguous}
% & %\multirow{3}{*}{31}
% & %\multirow{3}{*}{0.03\%}
% & \textsf{away} & 4 & 0.00\% \\
%& 31 & 0.03\% & \textsf{origin} & 15 & 0.01\% \\
%&&& \textsf{straddle} & 12 & 0.01\% \\ \hline
Affine, non-contiguous
 & 3
 & 0.00\% & \\ \hline
\multirow{3}{*}{Neighbour, contiguous}
 & %\multirow{3}{*}{83689}
 & %\multirow{3}{*}{83.08\%}
 & \textsf{away} & 152 & 0.14\% \\
& 83689 & 76.94\% & \textsf{origin} & 81309 & 74.75\% \\
&&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
\multirow{3}{*}{Neighbour, non-contiguous}
& & & \textsf{away}         & 2         & 0.00\%  \\
& 4854 & 4.46\% & \textsf{origin}       & 4836      & 4.45\%  \\
& & & \textsf{straddle}     & 16        & 0.01\%  \\ \hline
%Neighbour, non-contiguous
% & 4854
% & 4.46\% \\ \hline
Other & 20196 & 18.57\%  \\
 \hline \hline
\textbf{Total}  & 108773    & 100.00\% \\
\end{tabular}
\end{center}
%%
Thus, $76.94\%$ of array computations have only neighbour
indices (which may include some absolute indices) in a contiguous
pattern on the right hand side, with $76.80\%$ either including the origin
($74.75\%$) or offset by a distance of 1 from the origin
($2.05\%$). This confirms both Hypothesis 2 and Hypothesis 3. Only 152 instances
of index patterns not immediately next-to or over the origin were
found.  There is a non trivial amount of non-contiguous arrays access,
but it is a much smaller proportion of the whole, and mostly contains
the origin.

@@

We subclassified our data by RHS
and LHS categories

@@

neighbour-indexed or affine-indexed array. We additionally

@@

& \textbf{\# consistent}

@@

The most common category is an LHS array subscript

@@

Most stencils are contiguous, but there is some
non-contiguity. Most stencils have matching induction
variables on the left and right-hand sides but again there
are non-trivial amounts of the other kinds of relationship.
\item Many array computations read from each particular index just
  once.
We classified right-hand sides comprising neighbourhood (or
affine) subscripts and subclassified this by whether array subscript terms were
all unique or whether there were some repeats.
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  %\\\hline\hline
%Total                             & 108773    &
Thus most array computations have access patterns with
unique use of array subscripts, confirming Hypothesis 5.

@@

\noindent
For each of the $108,773$ array computations, we recorded three
additional properties: the dimensionality of array
subscripts involved, %the maximum depth of neighbourhood offsets,
the number of subscript terms contributing to the computation, and the
length of the dataflow path (\ie{}, how many intermediate assignments
contribute to the array computation):%The following gives the data
%as histograms:

@@

 \begin{center}
\begin{tabular}{l|ccccccccccccccccccccc}
  \textbf{Dataflow length}: & 0      & 1     & 2    & 3
  & 4    & 5 & 6    & 7   & 8   & 9   & 10  & 11  & 12  & 13  & 14
  & 15$\leq$ \\\hline
  Count & 106,712 & 13,323 & 7,091 & 3,898 & 3,031 & 1,695 & 1,333 & 965 & 792 & 704 & 482 & 430 & 339 & 310 & 270 & 2544
\end{tabular}
\end{center}
%
The data for ``number of subscripts'' shows that most computations are
relatively simple, with only 1-3 array subscripts.  However, some are
complex, with 184 with over 20 indexing terms. We had one example with
96 contributing array subscripts. Thus, we see that there are many
cases in practice with complex patterns. We aim to provide
support for such code with our specification language.

@@

\item Access patterns are mostly contiguous ($76.94\%$ of all array
  computations) and cross or straddle the origin ($75.75\%$), thus we
  can represent patterns by \emph{finite intervals} that meet at, or
  are next to, the origin.  These intervals will be given
  declaratively and abstractly, avoiding replicating indexing terms at
  the specification level.

@@

\item Stencils will be the primary focus $(61.79\%)$ but we will also
  allow specifications on assignments where the LHS is a variable
  (\eg{}, in a reduction) ($4.22\%$).

@@

\item Affine indexing is very rare $<0.03\%$ so we do not consider it.
\end{itemize}
